---
title: "Blake_Iter+Func"
author: "Kylie Blake"
date: "2025-03-26"
output:
  pdf_document:
    toc: true
  md_document:
    variant: gfm
---

1. Load in necessary packages 
```{r}
library(ggplot2)
library(drc)
library(tidyverse)
```

# Functions
```{r}
#Learning about functions
#(5*degree_f - 32)/9
#Calculating Celsius:
(5*(32 - 32)/9)
(5*(36 - 32)/9)
(5*(39 - 32)/9)
```
Rather than copy and pasting the same equation and changing the degree_f value, which can lead to typos, we can create a function instead. 

### Sample function

> sample.function <- function(... variable goes here ...){
  .... code goes here.... 
  return(... output ...)
} 

```{r}
# Name a function
F_to_C <- function(f_temp){
 celsius <- (5*(f_temp -32)/9)
 return(celsius)
} 
#in parentheses put the input for that function, it can be anything
#input f_temp into this caluculation, save that resulting value as Celsius, then return that value 

#To use the function:Enter a value you want converted from Fahrenheit to Celsius
F_to_C(32) 
F_to_C(90)
```
# Iterations
- Useful for reproducibility to prevent copy and paste errors

```{r}
#Repetition:
rep("A",3) 
rep(c("A","B"), 10)
rep(c(1,2,5,3), 4, each =5) #repeat sequence of numbers 4 times with each number repeated 5 times 
```
```{r}
#Sequence: sequence of #s

1:7 #this is a sequence, but use seq() as get more complicated 
seq(from=1, to =7)
seq(from= 0, to = 10, by=2)
```
```{r}
#Seq_along: allows you to generate sequence of numbers based on character vector 

LETTERS #gives alphabet of characters

seq_along(LETTERS) #counts number of characters as sequence 
```
# The for loop
**For loop:** sets up iteration of initialization 
Ask question: Is that sequence over? Will repeat 


**Example 1**

```{r}
for (i in 1:10) {
print(i*2)  
}
```
Note: Inside parentheses for for loop we define our iteration. Usually use an i as general syntax. Setting i = 1 through 10. Print ix2 so essentially 1 through 10 x2. You use print so you can see the result in the console. 

**Example 2**
```{r}
#Caluclate Fahrenheit for each value -30 through 100
for (i in -30:100){
  result <- F_to_C(i)
  print(result)
}

#To store values from for loop

# Initialize an empty data frame to store the results
celsius.df <- NULL

# Loop through the range of temperatures from -30 to 100
for (i in -30:100) {
  # Convert the current temperature from Fahrenheit to Celsius and store it in a data frame
  result <- data.frame(F_to_C(i), i)
  
  # Add the result to the celsius.df data frame
  celsius.df <- rbind.data.frame(celsius.df, result)
  
  # Print the current result
  print(result)
}

print(celsius.df)
```

