---
title: "DataWrangling"
author: "Kylie Blake"
date: "2025-03-17"
output: html_document:
  md_document:
    variant: gfm
  pdf_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
---

```{r}
library(tidyverse)
microbiome.fungi <- read.csv("CodingChallenges\\Bull_richness.csv")
str(microbiome.fungi) 
```
1. Select function 
#### "select()"
- You can use this function to select only certain columns of your data
```{r}
microbiome.fungi2 <- select(microbiome.fungi,SampleID,Crop, Compartment:Fungicide, richness)
#colon in btw comp and fungicide, can use this if have number of columns in order to select the columns "compartment through fungicide" 
```

2. Filter function
#### "filter()"
- To filter out data by specific variables 
```{r}
#essentially does same thing a filter

head(filter(microbiome.fungi2,Treatment == "Conv.")) #filter this by where treatment is equal to conventional 

#A more complex using &
head(filter(microbiome.fungi2, Treatment == "Conv." & Fungicide == "C")) #filtering by two conditions

#Another more complex example using or | 
head(filter(microbiome.fungi2, Sample == "A" | Sample == "B")) #samples A or B

```

3. Mutate function
#### "mutate()" 
- Allows us to quickly create new columns 

```{r}
microbiome.fungi2$logRich <- log(microbiome.fungi2$richness)  #is the same as:

#Create a new column called logRich
head(mutate(microbiome.fungi2, logRich = log(richness))) #format (dataset, name of column = function)

#Creating a new column which combines Crop and Treatment
head(mutate(microbiome.fungi2, Crop_Treatment = paste(Crop, Treatment))) 

```
4. The pipe 
#### "%>%" 
The pipe allows you to string multiple commands together, and helps to visualize what you are doing to your data step by step. 
- Allows you to input previous data from the right side of the pope into the function on the left (or next line) 

```{r}
#take orig dataset then the pipe: (do not need to define dataset in each command) 
microbiome.fungi %>% 
  select(SampleID, Crop, Compartment:Fungicide, richness) %>%  #selecting columns 
  filter(Treatment == "Conv.") %>%  #subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% #creating a new column of the log richness
  head() #display the first six rows 


```

5. Summarise 
#### "summarise()"
We can use this function to calculate mean, std dev and errors
```{r}
microbiome.fungi %>% 
  select(SampleID, Crop, Compartment:Fungicide, richness) %>%  #selecting columns 
  filter(Treatment == "Conv.") %>%  #subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% #creating a new column of the log richness
  summarise(Mean.rich = mean(logRich)) #calculating overall mean log richness withing the conventionally merged treatment 
#This will only give one calculation, which is the average log(richness) 

microbiome.fungi %>% 
  select(SampleID, Crop, Compartment:Fungicide, richness) %>%  #selecting columns 
  filter(Treatment == "Conv.") %>%  #subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% #creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), #calculating overall mean log richness withing the conventionally merged treatment 
            n=n(), #calculates the number of rows, give column name of n
            sd.dev = sd(logRich)) %>% #give column name sd.dev, calculate std.dev
  mutate(std.err = sd.dev/sqrt(n)) #mutate function to calculate std error 

```
6. Group-by
#### "group_by()"
Allows us to group data by a certain way 

```{r}
#CALUCLATE MEANS AND STD DEV FOR EACH TREATMENT AND FUNGICIDE
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n))
```

#### Connnecting to plot 
We can use pipes directly in a ggplot! 

```{r}

microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n)) %>%
  #Adding in ggplot - Do not need to specify the data set since you specified at the top 
  ggplot(aes(x = Fungicide, y = Mean.rich)) + # adding in a ggplot
  geom_bar(stat="identity") +
  geom_errorbar( aes(x=Fungicide, ymin=Mean.rich-std.err, ymax=Mean.rich+std.err), width=0.4) +
  theme_minimal() +
  xlab("") +
  ylab("Log Richness") +
  facet_wrap(~Treatment)
```
#### Joining 
Functions that allow you to combine multiple datasets based on a common set of variables 
- left_join()
  - keep all rows of X and add matching rows from Y. Any rows in Y that don't match X are excluded
-right_join()
  -reverse of left_join()
- inner_join()
  -only keep rows that are common to both X and Y, remove everything else
-full_join()
  - keep any columns that are in either X or Y
  
```{r}
# selecting just the richness and sample ID
richness <- microbiome.fungi %>%
  select(SampleID, richness)

# selecting columns that don't include the richness
metadata <- microbiome.fungi %>% 
  select(SampleID, Fungicide, Crop, Compartment, GrowthStage, Treatment, Rep, Sample)

head(metadata)
head(richness)

#Can join these two dataframes by common column == SampleID

head(left_join(metadata,richness, by= "SampleID")) #adding the richness data to the metadata based on the common column of sampleID 
```
  
#### Pivoting
long data -> one observation per sample or per row 
wide data -> multiple observations per row 
*wide-data is not wrong per say, but in general we want long-data
```{r}
#Set up data frame
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  filter(Class == "Sordariomycetes") %>%
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) # calculates the mean per Treatment and Fungicide 

#Change from long format (above) to wide format 
#Change dataset to this: two columns C anf F column for fungicide -- we want names from fungicide column to be put in new columns 
#values input to be taken from mean column
microbiome.fungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  filter(Class == "Sordariomycetes") %>%
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide 
  pivot_wider(names_from = Fungicide, values_from = Mean) # pivot to wide format
```

Now we can easily take the difference between the fungicide and control

```{r}
microbiome.fungi %>% 
  select(SampleID, Crop, Compartment:Fungicide, richness ) %>%  #selecting columns 
  group_by(Treatment, Fungicide) %>%  #grouping by treatment and fungicide to later calc summary stats by grp 
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide
  pivot_wider(names_from = Fungicide, values_from = Mean) %>%  #pivot to wide format
  mutate(diff.fungicide= C-F) # calculate the diff btw the means 
```

