---
title: "PLPA6820_Study"
author: "Kylie Blake"
date: "2025-03-04"
output: 
  html_document:
    md_document:
      variant: gfm
---

## Quiz Activity: R Programming Basics
Question 1: Subsetting Data
Using the subset function: How would you create a subset of the iris dataset where Species is "virginica"?
```{r}
data("iris")
subset(iris[iris$Species=="virginica",])
#can also do
subset(iris, Species == "virginica")
```


Without using the subset function: How would you create the same subset using indexing?
```{r}
subset <- (iris[iris$Species=="virginica",])
```

Question 2: Mathematical Functions
How would you calculate the median of Sepal.Width for the entire iris dataset?
```{r}
median(iris$Sepal.Width)
```


How would you find the variance of Petal.Length for the "setosa" species?
```{r}
setosa_subset <- subset(iris, Species == "setosa")
var(setosa_subset$Petal.Length)
```


Question 3: Data Visualization with ggplot2
How would you create a histogram of Sepal.Length using ggplot2?
```{r}
library(ggplot2)
ggplot(iris, aes(x=Sepal.Length))+
  geom_histogram()
```


How would you create a line plot of Petal.Width vs Petal.Length, colored by Species?
```{r}
ggplot(iris, aes(x=Petal.Length, y=Petal.Width, color=Species))+
  geom_smooth()
```
## Extended Quiz Activity: R Programming Basics
Question 1: Subsetting Data
Using the subset function: How would you create a subset of the iris dataset where Sepal.Length is greater than 6?
```{r}
subset_Sepal.Length <- subset(iris, Sepal.Length >6)
subset_Sepal.Length2 <- iris[iris$Sepal.Length>6,]
```

Without using the subset function: How would you create the same subset using indexing?
```{r}
subset_Sepal.Length2 <- iris[iris$Sepal.Length>6,]
```


Using the dplyr package: How would you create a subset of the iris dataset where Petal.Width is less than 1.5?
```{r}
library(dplyr)

iris %>%
  filter(Petal.Width<1.5)
```


Question 2: Mathematical Functions
How would you calculate the range of Sepal.Width for the entire iris dataset?
```{r}
range(iris$Sepal.Width)
```


How would you find the interquartile range (IQR) of Petal.Length for the "versicolor" species?
```{r}
subset_PL <- iris[iris$Species == "versicolor",]
IQR(subset_PL$Petal.Length)
```

How would you calculate the sum of Sepal.Length for the "setosa" species?
```{r}
subset_SL <- iris[iris$Species == "setosa",]
iris_SL <- subset(iris, Species == "setosa")
sum(subset_SL$Sepal.Length)
sum(iris_SL$Sepal.Length)
```

Question 3: Data Manipulation
Using the dplyr package: How would you add a new column to the iris dataset that contains the ratio of Sepal.Length to Sepal.Width?
```{r}
library(dplyr)
iris_new <- iris %>% 
  mutate(Sepal.Ratio = Sepal.Length/Sepal.Width)
  
```


Using the dplyr package: How would you group the iris dataset by Species and calculate the mean Petal.Length for each species?
```{r}
library(dplyr)

mean_PL <- iris %>% 
  group_by(Species) %>% 
  summarise(mean_PL = mean(Petal.Length))
```


Question 4: Data Visualization with ggplot2
How would you create a bar plot of the count of each Species using ggplot2?
```{r}
ggplot(iris, aes(x=Species, fill=Species))+
  geom_bar()+
  labs(x="Species", y="count")
```


How would you create a density plot of Sepal.Length, filled by Species?
```{r}
ggplot(iris, aes(x=Sepal.Length, fill = Species))+
  geom_density()
```


How would you create a faceted scatter plot of Petal.Length vs Petal.Width, with each facet representing a different Species?
```{r}
ggplot(iris, aes(x=Petal.Length, y=Petal.Width))+
  geom_point()+
  geom_jitter() +
  facet_wrap(~Species)
```


#Coding Challenge 1

1. Read the CSV file "TipsR.csv" into a dataframe named df, treating . as NA.
```{r}
df <- read.csv("TipsR.csv", na.strings =".")
```


Create a vector named z with the values 1 to 200.
```{r}
z <- 1:200
```


Print the mean of z on the console. Print the standard deviation of z on the console.
```{r}
mean(z)
sd(z)
```


Create a logical vector named zlog that is TRUE for z values greater than 30 and FALSE otherwise.
```{r}
zlog <- z>30
```



Make a dataframe with z and zlog as columns. Name the dataframe zdf.
```{r}
zdf <- data.frame(z,zlog)
```

Change Column Names:

Change the column names in your new dataframe to "zvec" and "zlogic".
```{r}
colnames(zdf) <- c("zvec","zlogic")
```


Make a new column in your dataframe equal to zvec squared (i.e., z^2). Call the new column zsquared.
```{r}
zdf$zsquared <- zdf$zvec^2
```


Subset the dataframe with and without the subset() function to only include values of zsquared greater than 10 and less than 100.
```{r}
subset(zdf, zsquared >10 & zsquared <100)
zdf[zdf$zsquared >10 & zdf$zsquared >100,]
```


Subset the zdf dataframe to only include the values on row 26.
```{r}
zdf[26,]
```

Subset the zdf dataframe to only include the values in the column zsquared in the 180th row.
```{r}
zdf$zsquared[180]
```
