---
title: "CodingChallenge5"
author: "Kylie Blake"
date: "2025-03-22"
output:
  md_document:
    variant: gfm
  pdf_document:
    toc: true
---

# Question 1

**1.	3 pts. Download two .csv files from Canvas called DiversityData.csv and Metadata.csv, and read them into R using relative file paths.**

```{r}

#Loading in necessary packages
library(tidyverse)
library(ggplot2)

#Loading in DiversityData and Metadata
Metadata <- read.csv("CodingChallenge5//Metadata.csv", na.strings = "na")
DiversityData <-  read.csv("CodingChallenge5//DiversityData.csv")
```

# Question 2

**2.	4 pts. Join the two dataframes together by the common column ‘Code’. Name the resulting dataframe alpha.**

```{r}
#Combining the two datasets by column "Code"
alpha <- left_join(Metadata,DiversityData, by = "Code")
head(alpha) #viewing first few rows of dataframe
```

# Question 3

**3.	4 pts. Calculate Pielou’s evenness index: Pielou’s evenness is an ecological parameter calculated by the Shannon diversity index (column Shannon) divided by the log of the richness column.**
**a.	Using mutate, create a new column to calculate Pielou’s evenness index.** 
**b.	Name the resulting dataframe alpha_even.**

```{r}
#Creating Pielous's evenness index column, named PEI
alpha_even <- mutate(alpha, PEI = shannon/log(richness))
head(alpha_even) #viewing first few rows of dataframe
```

# Question 4

**4.	4. Pts. Using tidyverse language of functions and the pipe, use the summarise function and tell me the mean and standard error evenness grouped by crop over time.**
**a.	Start with the alpha_even dataframe** 
**b.	Group the data: group the data by Crop and Time_Point.**
**c.	Summarize the data: Calculate the mean, count, standard deviation, and standard error for the even variable within each group.**
```{r}
#caluculate summary statistics for evenness grouped by crop over time
alpha_average <-  alpha_even %>% 
  group_by(Crop, Time_Point) %>% #viewing data grouped by crop and time_point
  summarise(mean.even = mean(PEI), #calculate mean evenness
            count = n(), #count number of values by crop over time
            sd.even = sd(PEI), #calculate std dev evenness
            se.even = sd.even/sqrt(count)) #calculate std error evenness
head(alpha_average) #view some of the resulting alpha_average dataframe
```
# Question 5
**5.	4. Pts. Calculate the difference between the soybean column, the soil column, and the difference between the cotton column and the soil column**
**a.	Start with the alpha_average dataframe**
**b.	Select relevant columns: select the columns Time_Point, Crop, and mean.even.**
**c.	Reshape the data: Use the pivot_wider function to transform the data from long to wide format, creating new columns for each Crop with values from mean.even.**
**d.	Calculate differences: Create new columns named diff.cotton.even and diff.soybean.even by calculating the difference between Soil and Cotton, and Soil and Soybean, respectively.**
**e.	Name the resulting dataframe alpha_average2**

```{r}
alpha_average2 <- alpha_average %>% 
  select(Time_Point, Crop, mean.even) %>% #only select the three columns (Time_Point, Crop, mean.even)
  pivot_wider(names_from = Crop, values_from = mean.even) %>%  #put multiple observations in one row for separate columns by Crop
  mutate(diff.cotton.even = Soil - Cotton,
         diff.soybean.even = Soil - Soybean) #create new columns 
head(alpha_average2) #view some of the resulting alpha_average2 dataframe
```

# Question 6
**6.	4 pts. Connecting it to plots**
**a.	Start with the alpha_average2 dataframe**
**b.	Select relevant columns: select the columns Time_Point, diff.cotton.even, and diff.soybean.even.**
**c.	Reshape the data: Use the pivot_longer function to transform the data from wide to long format, creating a new column named diff that contains the values from diff.cotton.even and diff.soybean.even.**
**i.	This might be challenging, so I’ll give you a break. The code is below.**
**d.	Create the plot: Use ggplot and geom_line() with ‘Time_Point’ on the x-axis, the column ‘values’ on the y-axis, and different colors for each ‘diff’ category. The column named ‘values’ come from the pivot_longer. The resulting plot should look like the one to the right.** 
```{r}
alpha_average2 %>% 
  select(Time_Point, diff.cotton.even,diff.soybean.even) %>%#selecting only relevant columns
  pivot_longer(c(diff.cotton.even, diff.soybean.even), names_to = "diff") %>% #putting data in long format 
  ggplot(aes(x = Time_Point, y = value, color = diff)) + # adding in a ggplot
  geom_line() + #make line bar
  xlab("Time (hrs)") + #label x-axis
  ylab("difference from soil in Pielou's evenness") #label y-axis
 
```

# Question 7

7.	2 pts. Commit and push a gfm .md file to GitHub inside a directory called Coding Challenge 5. Provide me a link to your github written as a clickable link in your .pdf or .docx

**[Click here to my GitHub](https://github.com/kzb0180)**
